---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Loading and preprocessing the data
```{r}
activity <- read.csv("activity.csv")
summary(activity)
```

## Mean total number of steps taken
```{r histogram}
steps_per_day <- tapply(activity$steps, activity$date, sum)
hist(steps_per_day, col = "blue", main = "Histogram of Total Steps Taken per Day",
     xlab = "Total Steps Taken per Day")
```

```{r}
options(digits = 6)
org_mean <- mean(steps_per_day, na.rm = TRUE)
org_median <- median(steps_per_day, na.rm = TRUE)
```
The estimated mean is `r org_mean`, while the estimated median is `r org_median`.

## Average daily activity pattern
Time series plot of the average number of steps taken (averaged across all days).
```{r}
int <- unique(activity$interval)
avg_steps_across_days <- tapply(activity$steps, activity$interval, mean, na.rm = TRUE)
plot(int, avg_steps_across_days, type = "l", main = "Time series of average number of steps (averaged across all days)", xlab = "5-minute intervals", ylab = "Average number of steps")
```

Determine the 5-minute interval that contains the maximum number of steps.
```{r}
max_steps <- max(avg_steps_across_days)
index_max <- which(avg_steps_across_days == max_steps)
max_steps_int <- activity$interval[index_max]
```
The 5-minute interval that contains the maximum number of steps is *`r max_steps_int`* (across all the days in the dataset).

## Imputing missing values
The total number of missing values in the dataset (i.e. the total number of rows with NAs).
```{r}
sum(is.na(activity$steps))
```

A new dataset with the NAs replaced by the average number of steps for that given interval.
```{r}
activity_NA_rem <- activity
index_NAs <- which(is.na(activity_NA_rem$steps))
int <- unique(activity$interval)

for (i in index_NAs){
    interval_of_i <- activity_NA_rem$interval[i]
    activity_NA_rem$steps[i] <- avg_steps_across_days[which(int == interval_of_i)]
}
sum(is.na(activity_NA_rem$steps))
```

A histogram of the total number of steps taken each day.
```{r}
steps_per_day_NA_rem <- tapply(activity_NA_rem$steps, activity_NA_rem$date, sum)
hist(steps_per_day_NA_rem, col = "blue", main = "Histogram of total steps taken per Day",
     xlab = "Total steps taken per day")
```
 
Mean and median of dataset where NAs have been replaced.
```{r}
options(digits = 6)
imput_mean <- mean(steps_per_day_NA_rem)
imput_median <- median(steps_per_day_NA_rem)
```
The mean with imputted values is, `r imput_mean`. The median with the imputted values is `r imput_median`.

```{r}
diff_mean <- abs(org_mean - imput_mean)
diff_median <- abs(org_median - imput_median)
```

When NA values were removed the mean and median with imputted values are, `r diff_mean` and `r diff_median`, respectively. As we can see, there is no difference between the mean, and a slight difference with the median.

## Comparison in activity patterns between weekdays and weekends

Using the dataset with the filled-in missing values for this part a new factor variable in the dataset with two levels is created -- "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.
```{r}
activity_NA_rem$day <- weekdays(as.Date(activity_NA_rem$date))
#set all values to "weekday"
activity_NA_rem["dayType"] <- "weekday"
#change Saturday & Sundy values to "weekend"
activity_NA_rem$dayType[activity_NA_rem$day %in% c("Saturday", "Sunday")] <- "weekend"
```

A panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).
```{r}
avg_steps_df <- tapply(activity_NA_rem$steps, list(activity_NA_rem$interval, activity_NA_rem$dayType), mean)
avg_steps_df <- data.frame(avg_steps_df)
```
```{r}
par(mar = c(2,2,2,1), mfrow = c(2,1), oma=c(2.5,2.5,1,1))
plot(int, avg_steps_df$weekday, type = "l", xlab = "weekday", ylab = "", main = "Time Series Plot of Average Steps on the Weekend and Weekdays  ")
plot(int, avg_steps_df$weekend, type = "l", xlab = "weekend", ylab = "")
mtext(outer=T, side=1, line=1, text="Interval")  #caption for x-axis
mtext(outer=T, side=2, line=1, text="Number of steps taken")  #caption for y-axis
```

